Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.

### 2024-03-01 apples_and_oranges.qmd
Running some baseline analyses

@@

<!--# Charlotte, I am particularly keen for you to code independently the SE calculation, maybe use all the other parts as they are, but check the SEs. Once you have re-assured yourself about it, could you please read the code for the other steps below, but break it down by taking it out of the function and probably out of the loop and verifying that you would get the same results for one dataset, e.g. the first one in the list of 1000. It would be amazing to verify the results this way. Finally, as discussed, we can vary the parammeters of the distribution if we have better a prioris-->

@@

###########4. Function to extract the SMDs  (SMD, ses, CI) for each level of the dummy variable. The output is used below in 5.
# it is a slightly awkward one, because I couldn't' come up with a better way to add the intercept to each coefficient
# given the named output of the built in coef function. 
# here I extract the coefficients
extract_coefficients_func <- function(df_with_coefs){
  list_coefs <- list()
  coefs <- 0
  coefficient_output <- 0

@@

# Double coding SE for argyris
double_check <- list_df_simulated[[1]]

double_check <- double_check %>% 
  mutate (se_check = sqrt(  ((2*(1 - correlation_sim_values )/baseline_n) + (cohens_d^2/ 2*baseline_n) )))

view(double_check)

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
library(tidyverse)
# Start here------------------------------------------------------
# import master dataset

@@

### A few more tidying things from Argyris before doing metanalyses
# Discovered some errors in the percentage women variable. Have gone back and checked. 
df_appl_v_orange[df_appl_v_orange$study_ID=="Fristad, 2019_cbt + placebo_placebo",]$percent_women <-43.1

@@

# now just get rid of the unnecessary columns as these are causing clutter

@@

# save this as the master dataframe  --------------------------------------
# use this dataframe to perform all operations below in quarto

@@

Fourth, we tested whether our results could be accounted for by differential regression to the mean. In an initial step, we tested in metaregression as above for differences in baseline means. For this we used a min-max normalisation to turn each study arm mean score into a percentage using the following formalism:

$$
\text{outcome}_{percent} = \frac{\bar{X} - \text{scale}_{min}}{\text{scale}_{max} - \text{scale}_{min}}
$$ {#eq-14}

where, $$\bar{X}$$ is the mean score of each study arm on the primary questionnaire, and ${scale}_{min}$ and ${scale}_{max}$ are the minimum and maximum possible value of the scale in question, and for which the the standard deviation is calculated thus:

$$
\text{SD}_{Xpercent} = \frac{\text{SD}_{X}}{\text{scale}_{max} - \text{scale}_{min}}
$$ {#eq-15}

where $\text{SD}_{X}$ is the original standard deviation of the mean for X. Applying equations (12) and (13) on the baseline and outcome means and standard deviations of each study allowed us to then calculate SMDs and standard errors using equations (4) and (5)

@@

Studies included in the metaregression.

@@

@@

@@

### Analysis of all studies

#### Number and type of of studies included

#### Metaregression analysis.

@@

### do ALL studies using percentage at baseline
all_studies_baseline_metareg <- list_df_simulated_percentage[[1]]

#turn sds to sec
all_studies_baseline_metareg$baseline_stand_error <- all_studies_baseline_metareg$perc_baseline_sd/
                                        sqrt(all_studies_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
# also change this for the function to work
all_studies_baseline_metareg$baseline_mean <- all_studies_baseline_metareg$perc_baseline_mean
results_all_studies_metareg_baseline <- run_metareg_and_extract(all_studies_baseline_metareg,  
                                                         model_1, condition)

### get CDRS studies
cdrs_studies_for_baseline_metareg <- list_cdrs_study[[1]] # created this above, you can take any of these, as the means and sds are repeated.

#turn sds to sec
cdrs_studies_for_baseline_metareg$baseline_stand_error <- cdrs_studies_for_baseline_metareg$baseline_sd/
                                        sqrt(cdrs_studies_for_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout

results_cdrs_metareg_baseline <- run_metareg_and_extract(cdrs_studies_for_baseline_metareg,  
                                                         model_1, condition)

### get HAM-D studies
hamd_study_ids <-
  df_appl_v_orange[df_appl_v_orange$instrument_name == "hamd",]$new_study_id

list_hamd_study <- list_df_simulated[[1]][list_df_simulated[[1]]$new_study_id
                                                          %in% hamd_study_ids,]

hamd_studies_for_baseline_metareg <- list_hamd_study #

#turn sds to sec
hamd_studies_for_baseline_metareg$baseline_stand_error <- hamd_studies_for_baseline_metareg$baseline_sd/
                                        sqrt(hamd_studies_for_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
results_hamd_metareg_baseline <- run_metareg_and_extract(hamd_studies_for_baseline_metareg, 
                                                         model_1, condition)

results_hamd_metareg_baseline$condition <-  levels(list_df_simulated[[1]]$four_level_var)

# Now we want to do look at baseline severity when we exclude psy studies with sub clinical depression levels

# We will do in two ways
# First by excluding studies with a diagnostic classification of "sub"

clin_studies_baseline_metareg <- all_studies_baseline_metareg %>%
  filter(new_study_id %in% clin_study_ids)

#turn sds to sec
clin_studies_baseline_metareg$baseline_stand_error <- clin_studies_baseline_metareg$perc_baseline_sd/
                                        sqrt(clin_studies_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
# also change this for the function to work
clin_studies_baseline_metareg$baseline_mean <- clin_studies_baseline_metareg$perc_baseline_mean
results_clin_studies_baseline_metareg <- run_metareg_and_extract(clin_studies_baseline_metareg,  
                                                         model_1, condition)

# Now lets include only studies with classification of "mdd"

mdd_study_ids <-
  df_appl_v_orange[df_appl_v_orange$diagnosis == "mdd",]$new_study_id

mdd_studies_baseline_metareg <- all_studies_baseline_metareg %>%
  filter(new_study_id %in% mdd_study_ids)

mdd_studies_baseline_metareg %>% 
  filter(four_level_var %in% c("psychotherapy_active","psychotherapy_control")) %>%
  summarise(n = n())

#turn sds to sec
mdd_studies_baseline_metareg$baseline_stand_error <- mdd_studies_baseline_metareg$perc_baseline_sd/
                                        sqrt(mdd_studies_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
# also change this for the function to work
mdd_studies_baseline_metareg$baseline_mean <- mdd_studies_baseline_metareg$perc_baseline_mean
results_mdd_studies_baseline_metareg <- run_metareg_and_extract(mdd_studies_baseline_metareg,  
                                                         model_1, condition)

# Now CBT + all med studies

cbt_meds_studies_baseline_metareg <- all_studies_baseline_metareg %>%
  filter(new_study_id %in% cbt_meds_study_ids) %>% 
  filter(!(arm_effect_size == "cohens_d_active" & psy_or_med == 1 & !(type == "cbt")))

#turn sds to sec
cbt_meds_studies_baseline_metareg$baseline_stand_error <- cbt_meds_studies_baseline_metareg$perc_baseline_sd/
                                        sqrt(cbt_meds_studies_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
# also change this for the function to work
cbt_meds_studies_baseline_metareg$baseline_mean <- cbt_meds_studies_baseline_metareg$perc_baseline_mean
results_cbt_meds_studies_baseline_metareg <- run_metareg_and_extract(cbt_meds_studies_baseline_metareg,  
                                                         model_1, condition)

# Look at the diagnostics of included studies

filtered_data <- df_appl_v_orange %>%
  filter(psy_or_med == 1) %>% 
  filter(new_study_id %in% cbt_meds_study_ids) %>%
  distinct(new_study_id, .keep_all = TRUE)

summary_result <- filtered_data %>%
  group_by(diagnosis) %>%
  summarise(count = n())

# Now the CBT + med studies, filtering out the subclinical studies

clin_cbt_meds_study_ids <-
  df_appl_v_orange[(df_appl_v_orange$active_type == "cbt" | df_appl_v_orange$psy_or_med == 0) & df_appl_v_orange$diagnosis != "sub", ]$new_study_id

clin_cbt_meds_studies_baseline_metareg <- all_studies_baseline_metareg %>%
  filter(new_study_id %in% clin_cbt_meds_study_ids) %>% 
  filter(!(arm_effect_size == "cohens_d_active" & psy_or_med == 1 & !(type == "cbt")))

#turn sds to sec
clin_cbt_meds_studies_baseline_metareg$baseline_stand_error <- clin_cbt_meds_studies_baseline_metareg$perc_baseline_sd/
                                        sqrt(clin_cbt_meds_studies_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
# also change this for the function to work
clin_cbt_meds_studies_baseline_metareg$baseline_mean <- clin_cbt_meds_studies_baseline_metareg$perc_baseline_mean
results_clin_cbt_meds_studies_baseline_metareg <- run_metareg_and_extract(clin_cbt_meds_studies_baseline_metareg,  
                                                         model_1, condition)

# Looking at demographics at baseline
# Currently we have both control and active conditions, we just need each study once

met_perc_women <- metagen(TE = percent_women,
                           seTE = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)
met_perc_women

# ### get CDRS clin studies
#  clin_cdrs_studies_baseline_metareg <- cdrs_studies_for_baseline_metareg %>% 
#    filter(new_study_id %in% clin_study_ids)
#  
# 
# model_1 <- as.formula(~ four_level_var)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# 
# results_clin_cdrs_metareg_baseline <- run_metareg_and_extract(clin_cdrs_studies_baseline_metareg,  
#                                                          model_1, condition)
# 
# # Just realised this is identical, as all the studies that use the CDRS have clinical samples. 
# 
# ### get HAM-D clin studies
#  clin_hamd_studies_for_baseline_metareg <- hamd_studies_for_baseline_metareg %>% 
#    filter(new_study_id %in% clin_study_ids)
# 
# 
# model_1 <- as.formula(~ four_level_var)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# results_clin_hamd_studies_for_baseline_metareg <- run_metareg_and_extract(clin_hamd_studies_for_baseline_metareg, 
#                                                          model_1, condition)
# 
# results_clin_hamd_studies_for_baseline_metareg$condition <-  levels(list_df_simulated[[1]]$four_level_var)
# 
#

>>>>

### Sample characteristics

We will test in a metaregression for differences between psychotherapy versus medication trials in sample characteristics at baseline.

First, we compared depression severity levels at baseline. In order to compare across different instruments, we performed a min-max normalisation to turn each study arm mean score at baseline into a percentage using the following formalism:

$$
\text{outcome}_{percent} = \frac{\bar{X} - \text{scale}_{min}}{\text{scale}_{max} - \text{scale}_{min}}
$$ {#eq-3}

where, $$\bar{X}$$ is the mean score for each study arm on the primary outcome questionnaire, and ${scale}_{min}$ and ${scale}_{max}$ are the minimum and maximum possible values of the scale in question, respectively. The standard deviation is calculated thus:

$$
\text{SD}_{Xpercent} = \frac{\text{SD}_{X}}{\text{scale}_{max} - \text{scale}_{min}}
$$ {#eq-4}

where $\text{SD}_{X}$ is the original standard deviation of the mean at baseline.

We also conducted metaregressions to compare psychotherapy vs medication trials on the gender composition of the sample, mean age, and number of trial sites.

@@

@@

###########4. Function to extract the SMDs  (SMD, ses, CI) for each level of the dummy variable. The output is used below in 5.
# it is a slightly awkward one, because I couldn't' come up with a better way to add the intercept to each coefficient
# given the named output of the built in coef function. here I extract the coefficients

extract_coefficients_func <- function(df_with_coefs){
  list_coefs <- list()
  coefs <- 0
  coefficient_output <- 0

@@

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
library(tidyverse)
# Transforming data ------------------------------------------------------
# import master dataset

@@

### A few more tidying things from Argyris before doing metanalyses
# Discovered some errors in the percentage women variable. Have gone back and checked. 
df_appl_v_orange[df_appl_v_orange$study_ID=="Fristad, 2019_cbt + placebo_placebo",]$percent_women <-43.1
df_appl_v_orange[df_appl_v_orange$study=="Clarke, 1995",]$percent_women <-47.3
df_appl_v_orange[df_appl_v_orange$study=="Moeini, 2019",]$percent_women <-100
df_appl_v_orange[df_appl_v_orange$study=="Santomauro, 2016",]$percent_women <-40

@@

# Now just get rid of the unnecessary columns as these are causing clutter

@@

# save this as the master dataframe  --------------------------------------
# use this dataframe to perform all operations below

@@

@@

## Included studies

@@

## Sample characteristics at baseline in medication versus psychotherapy trials

Using a random effects meta-regression we compared psychotherapy versus medication trials on four domains: gender composition, age, baseline depression severity levels, and number of trial sites. We first examined each of these across all studies, and then conducted two sensitivity analyses where we excluded psychotherapy studies which 1) recruited participants with sub clinical depression levels and 2) did not use CBT as the active intervention.

### Gender

Gender composition was similar across psychotherapy and medication trials,

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# Start by looking at demographics at baseline
# Currently we have both control and active conditions, we just need each study once

# The metagen function doesn't like when NAs result in the TE vector being a different length to subgroup vector. Need to start by removing NAs

met_perc_women <- metagen(TE.random = percent_women,
                           seTE.random = percent_women_std_error,
                           studlab = new_study_id,
                           data = df_for_gender_metan,
                           sm = "",
                           fixed = FALSE,
                           random = TRUE,
                           method.tau = "REML",
                           hakn = TRUE,
                           title = "percentage women across studies",
                           subgroup = psy_or_med,
                           tau.common = FALSE)

summary_results <- summary(met_perc_women, byvar = "subgroup")

# Convert the results to a dataframe
result_df <- data.frame(
  TE = summary_met_perc_women$TE.random,
  seTE = summary_met_perc_women$seTE.random,
  lower = summary_met_perc_women$lower.random,
  upper = summary_met_perc_women$upper.random
)

knitr:: kable(summary_table, "latex", caption = "Gender composition")

# ### do ALL studies using percentage at baseline
# all_studies_baseline_metareg <- list_df_simulated_percentage[[1]]
# #turn sds to sec
# all_studies_baseline_metareg$baseline_stand_error <- all_studies_baseline_metareg$perc_baseline_sd/
#                                         sqrt(all_studies_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# # also change this for the function to work
# all_studies_baseline_metareg$baseline_mean <- all_studies_baseline_metareg$perc_baseline_mean
# results_all_studies_metareg_baseline <- run_metareg_and_extract(all_studies_baseline_metareg,  
#                                                          model_1, condition)
# ### get CDRS studies
# cdrs_studies_for_baseline_metareg <- list_cdrs_study[[1]] # created this above, you can take any of these, as the means and sds are repeated.
# #turn sds to sec
# cdrs_studies_for_baseline_metareg$baseline_stand_error <- cdrs_studies_for_baseline_metareg$baseline_sd/
#                                         sqrt(cdrs_studies_for_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# results_cdrs_metareg_baseline <- run_metareg_and_extract(cdrs_studies_for_baseline_metareg,  
#                                                          model_1, condition)
# ### get HAM-D studies
# hamd_study_ids <-
#   df_appl_v_orange[df_appl_v_orange$instrument_name == "hamd",]$new_study_id
# list_hamd_study <- list_df_simulated[[1]][list_df_simulated[[1]]$new_study_id
#                                                           %in% hamd_study_ids,]
# hamd_studies_for_baseline_metareg <- list_hamd_study # 
# #turn sds to sec
# hamd_studies_for_baseline_metareg$baseline_stand_error <- hamd_studies_for_baseline_metareg$baseline_sd/
#                                         sqrt(hamd_studies_for_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# results_hamd_metareg_baseline <- run_metareg_and_extract(hamd_studies_for_baseline_metareg, 
#                                                          model_1, condition)
# results_hamd_metareg_baseline$condition <-  levels(list_df_simulated[[1]]$four_level_var)
# # Now we want to do look at baseline severity when we exclude psy studies with sub clinical depression levels
# # We will do in two ways
# # First by excluding studies with a diagnostic classification of "sub"
# clin_study_ids <-
#   df_appl_v_orange[df_appl_v_orange$diagnosis != "sub",]$new_study_id
# clin_studies_baseline_metareg <- all_studies_baseline_metareg %>%
#   filter(new_study_id %in% clin_study_ids)
# #turn sds to sec
# clin_studies_baseline_metareg$baseline_stand_error <- clin_studies_baseline_metareg$perc_baseline_sd/
#                                         sqrt(clin_studies_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# # also change this for the function to work
# clin_studies_baseline_metareg$baseline_mean <- clin_studies_baseline_metareg$perc_baseline_mean
# results_clin_studies_baseline_metareg <- run_metareg_and_extract(clin_studies_baseline_metareg,  
#                                                          model_1, condition)
# # Now lets include only studies with classification of "mdd"
# mdd_study_ids <-
#   df_appl_v_orange[df_appl_v_orange$diagnosis == "mdd",]$new_study_id
# mdd_studies_baseline_metareg <- all_studies_baseline_metareg %>%
#   filter(new_study_id %in% mdd_study_ids)
# mdd_studies_baseline_metareg %>% 
#   filter(four_level_var %in% c("psychotherapy_active","psychotherapy_control")) %>%
#   summarise(n = n())
# #turn sds to sec
# mdd_studies_baseline_metareg$baseline_stand_error <- mdd_studies_baseline_metareg$perc_baseline_sd/
#                                         sqrt(mdd_studies_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# # also change this for the function to work
# mdd_studies_baseline_metareg$baseline_mean <- mdd_studies_baseline_metareg$perc_baseline_mean
# results_mdd_studies_baseline_metareg <- run_metareg_and_extract(mdd_studies_baseline_metareg,  
#                                                          model_1, condition)
# # Now CBT + all med studies
# cbt_meds_study_ids <-
#   df_appl_v_orange[df_appl_v_orange$active_type == "cbt" | df_appl_v_orange$psy_or_med == 0, ]$new_study_id
# cbt_meds_studies_baseline_metareg <- all_studies_baseline_metareg %>%
#   filter(new_study_id %in% cbt_meds_study_ids) %>% 
#   filter(!(arm_effect_size == "cohens_d_active" & psy_or_med == 1 & !(type == "cbt")))
# #turn sds to sec
# cbt_meds_studies_baseline_metareg$baseline_stand_error <- cbt_meds_studies_baseline_metareg$perc_baseline_sd/
#                                         sqrt(cbt_meds_studies_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# # also change this for the function to work
# cbt_meds_studies_baseline_metareg$baseline_mean <- cbt_meds_studies_baseline_metareg$perc_baseline_mean
# results_cbt_meds_studies_baseline_metareg <- run_metareg_and_extract(cbt_meds_studies_baseline_metareg,  
#                                                          model_1, condition)
# # Look at the diagnostics of included studies 
# filtered_data <- df_appl_v_orange %>%
#   filter(psy_or_med == 1) %>% 
#   filter(new_study_id %in% cbt_meds_study_ids) %>%
#   distinct(new_study_id, .keep_all = TRUE)
# summary_result <- filtered_data %>%
#   group_by(diagnosis) %>%
#   summarise(count = n())
# # Now the CBT + med studies, filtering out the subclinical studies
# clin_cbt_meds_study_ids <-
#   df_appl_v_orange[(df_appl_v_orange$active_type == "cbt" | df_appl_v_orange$psy_or_med == 0) & df_appl_v_orange$diagnosis != "sub", ]$new_study_id
# clin_cbt_meds_studies_baseline_metareg <- all_studies_baseline_metareg %>%
#   filter(new_study_id %in% clin_cbt_meds_study_ids) %>% 
#   filter(!(arm_effect_size == "cohens_d_active" & psy_or_med == 1 & !(type == "cbt")))
# #turn sds to sec
# clin_cbt_meds_studies_baseline_metareg$baseline_stand_error <- clin_cbt_meds_studies_baseline_metareg$perc_baseline_sd/
#                                         sqrt(clin_cbt_meds_studies_baseline_metareg$baseline_n)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# # also change this for the function to work
# clin_cbt_meds_studies_baseline_metareg$baseline_mean <- clin_cbt_meds_studies_baseline_metareg$perc_baseline_mean
# results_clin_cbt_meds_studies_baseline_metareg <- run_metareg_and_extract(clin_cbt_meds_studies_baseline_metareg,  
#                                                          model_1, condition)

# ### get CDRS clin studies
#  clin_cdrs_studies_baseline_metareg <- cdrs_studies_for_baseline_metareg %>% 
#    filter(new_study_id %in% clin_study_ids)
#  
# 
# model_1 <- as.formula(~ four_level_var)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# 
# results_clin_cdrs_metareg_baseline <- run_metareg_and_extract(clin_cdrs_studies_baseline_metareg,  
#                                                          model_1, condition)
# 
# # Just realised this is identical, as all the studies that use the CDRS have clinical samples. 
# 
# ### get HAM-D clin studies
#  clin_hamd_studies_for_baseline_metareg <- hamd_studies_for_baseline_metareg %>% 
#    filter(new_study_id %in% clin_study_ids)
# 
# 
# model_1 <- as.formula(~ four_level_var)
# condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
#                                                               #throughout
# results_clin_hamd_studies_for_baseline_metareg <- run_metareg_and_extract(clin_hamd_studies_for_baseline_metareg, 
#                                                          model_1, condition)
# 
# results_clin_hamd_studies_for_baseline_metareg$condition <-  levels(list_df_simulated[[1]]$four_level_var)
# 
#
```

@@

list_df_simulated_percentage[[1]]

list_df_simulated[[1]]

@@

#### Mean efficacy of psychotherapy versus medication control conditions

#### Quality and nature of psychotherapy controls across trials

@@
